---
title: "CPBS 7630 Module 2B - Dimensionality Reduction"
date: "February 18, 2016"
output: html_document
---

---

#### Contents:

* [Get cleaned data](#synapse)
* [Principal components analysis](#pca)
* [Evaluation](#evaluation)

---

<a name="synapse"/>

### Get the cleaned data from Synapse

Get the drug sensitivity data from [Costello, et al. A Community Effort to Assess and Improve Drug Sensitivity Prediction Algorithms (2014)](http://www.nature.com/nbt/journal/v32/n12/full/nbt.2877.html). 

```{r, echo = F, message = F}
library(synapseClient)

# Log in to Synapse
#synapseLogin(username = 'myusername', password = 'mypassword')
```

```{r, message = F}

# ------------------------------- GET DATA -------------------------------------
# Get cleaned, log-transformed gene expression and response data from Synapse
DREAM7.data <- synGet('syn5663580')
local.file.path <- DREAM7.data@filePath

# Load the RData file
load(local.file.path)
rm(local.file.path, DREAM7.data)

# Enviroment now includes:
#############################
# expression.data - matrix of gene expression values for 30,154 genes in 39 cell lines
# drug7.response - response vector for Drug 7
# genes - vector of original gene names
# cell.lines - vector of 39 cell lines with both expression and response data
# all.response - response matrix for all drugs
#############################

# ------------------------------- FEATURE SELECTION ---------------------------

# For scaling for PCA later, we can't have any all-zero genes
expression.data <- data.frame(expression.data)
expression.data <- expression.data[, colSums(expression.data) > 0]

```

<a name="pca"/>

### Principal components analysis

Principal components analysis (PCA) is used for performing dimensionality reduction (feature extraction) to find novel, meaningful patterns in a data set.

#### Example 1 - toy data

To illustrate the basic principles of PCA, we will first look at a toy data set and an example adapted from [here](https://tgmstat.wordpress.com/2013/11/28/computing-and-visualizing-pca-in-r/#ref1). The `iris` data set contains data on 3 species of iris and measurements for their sepal length, sepal width, petal length and petal width.

```{r, message = F, warning = F}

# Toy data set
data(iris)

# Log transform 
iris.log <- log(iris[, 1:4])
iris.species <- iris[, 5]
 
# Apply PCA with stats::prcomp
iris.pca <- prcomp(iris.log,
                 center = TRUE,
                 scale. = TRUE)
# View the principal components rotation (also called loadings)
iris.pca

# Get ggbiplot from Github if needed
# library(devtools)
# install_github("vqv/ggbiplot")

# Plot principal component variance
plot(iris.pca, type = "l", col = 'skyblue', pch = 19, 
     main = 'PC variance for iris data')

# Plot PCA with ggbiplot 
library(ggbiplot)
g <- ggbiplot(iris.pca, obs.scale = 1, var.scale = 1, 
              groups = iris.species, ellipse = TRUE, 
              circle = TRUE)
g <- g + scale_color_discrete(name = '')
g <- g + theme(legend.direction = 'horizontal', 
               legend.position = 'top')
print(g)

# Summarize each of the principal components
summary(iris.pca)

```

#### Example 2 - PCA for drug synergy

In this example, we'll use the DREAM7 drug synergy data we prepared in the first code chunk.

```{r, message = F, warning = F}

# Apply PCA with stats::prcomp
expression.pca <- prcomp(expression.data,
                 center = TRUE,
                 scale. = TRUE)

# View the number of principal components
length(expression.pca$sdev)

# Plot principal component variance
plot(expression.pca, type = "l", col = 'skyblue', pch = 19, 
     main = 'PC variance for iris data')

# Get ggbiplot from Github if needed
# library(devtools)
# install_github("vqv/ggbiplot")

# Categorize drug response for coloring the plot
# Less than median GI50 = sensitive; greater than median = resistant
response.groups <- factor(drug7.response >= median(drug7.response), labels = c('sensitive', 'resistant'))

# Plot PCA with ggbiplot 
library(ggbiplot)
g <- ggbiplot(expression.pca, obs.scale = 1, var.scale = 1, 
              groups = response.groups, ellipse = TRUE, 
              var.axes = F, circle = TRUE)
g <- g + scale_color_discrete(name = '')
g <- g + theme(legend.direction = 'horizontal', 
               legend.position = 'top')
print(g)

# We could also try separating the groups more based on response distribution
hist(drug7.response, xlab = 'GI50')

response.groups <- factor(cut(drug7.response, 
                       breaks = c(0, 4.2, 6.7, 7.1),
                       labels = c('sensitive', 'average', 'resistant')))

g <- ggbiplot(expression.pca, obs.scale = 1, var.scale = 1, 
              groups = response.groups, ellipse = TRUE, 
              var.axes = F, circle = TRUE)
g <- g + scale_color_discrete(name = '')
g <- g + theme(legend.direction = 'horizontal', 
               legend.position = 'top')
print(g)

# Summarize each of the 39 principal components
summary(expression.pca)
```
  
<a name="evaluation"/>

### Comparison with/without PCA

```{r warning = F}

# Mean squared error
MSE <- function(yhat, y){
    squared.error <- sum(mapply(function(yhat,y) (yhat-y)^2, yhat, y))
    return(squared.error/length(yhat))
}

# Make train and test sets
train = cell.lines %in% sample(cell.lines, 20)
test = !train

# Feature matrix made from 29 random genes
old.feature.matrix <- expression.data[,sample(29)]
old.combined <- cbind(old.feature.matrix, Response = drug7.response)
old.lm <- lm(Response ~ ., old.combined[train,])
MSE(predict(old.lm, old.combined[test,], type="response"), old.combined[test, 'Response'])

# New feature matrix made from 29 PCs
new.feature.matrix <- data.frame(expression.pca$x[,1:29])
new.combined <- cbind(new.feature.matrix, Response = drug7.response)
new.lm <- lm(Response ~ ., new.combined[train,])
MSE(predict(new.lm, new.combined[test,], type="response"), new.combined[test, 'Response'])

```

#### TODO

Apply PCA to your new data set for the presentation on Thursday. How many principal components are required to explain 90% of the variance? Evaluate the effect of performing PCA on your results. Can you relate/correlate any of the principal components with the underlying biology (or other applicable data type)?